---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# decoupleR

<!-- badges: start -->
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![BioC status](http://www.bioconductor.org/shields/build/release/bioc/decoupleR.svg)](https://bioconductor.org/checkResults/release/bioc-LATEST/decoupleR)
[![BioC dev status](http://www.bioconductor.org/shields/build/devel/bioc/decoupleR.svg)](https://bioconductor.org/checkResults/devel/bioc-LATEST/decoupleR)
[![R build status](https://github.com/saezlab/decoupleR/workflows/R-CMD-check-bioc/badge.svg)](https://github.com/saezlab/decoupleR/actions)
[![Codecov test coverage](https://codecov.io/gh/saezlab/decoupleR/branch/master/graph/badge.svg)](https://codecov.io/gh/saezlab/decoupleR?branch=master)
[![GitHub issues](https://img.shields.io/github/issues/saezlab/decoupleR)](https://github.com/saezlab/decoupleR/issues)
<!-- badges: end -->

## Overview

Computational methods allow the extraction of mechanistic signatures from omics data based on prior knowledge resources, reducing the dimensionality of the data for increased statistical power and better interpretability. Here, we present decoupleR, a Bioconductor package containing different statistical methods to extract these signatures within a unified framework. decoupleR allows the user to flexibly test any method with any resource. It incorporates methods that take into account the sign and weight of network interactions.

For more information about how this package has been used with real data,
please check the following links:  

- [Decoupling statistics and networks: a crowdsourced systematic assessment of transcription factor activity estimation from transcriptomics data](https://docs.google.com/document/d/1aYD3IF1tj7H6vUFqKnfHB6d6Trs_oit7umjk5mQDNZ0/edit).
- [Creation of benchmarking pipelines](https://github.com/saezlab/decoupleR/tree/benchmark).

## Installation instructions

Get the latest stable `R` release from
[CRAN](http://cran.r-project.org/). 

Then install `decoupleR` using from [Bioconductor](http://bioconductor.org/) the
following code:

```{r bioconductor_install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("decoupleR")

# Check that you have a valid Bioconductor installation
BiocManager::valid()
```

Then install development version from [GitHub](https://github.com/) with:

```{r github_installation, eval = FALSE}
BiocManager::install("saezlab/decoupleR")
```

## Usage

### Load packaga and data

```{r usage-load_data, message=FALSE}
library(decoupleR)

inputs_dir <- system.file("testdata", "inputs", package = "decoupleR")

mat <- file.path(inputs_dir, "input-expr_matrix.rds") %>%
    readRDS() %>%
    dplyr::glimpse()

network <- file.path(inputs_dir, "input-dorothea_genesets.rds") %>%
    readRDS() %>%
    dplyr::glimpse()
```

### Decouple wrapper

`decouple()` allows access to all **decoupleR** available statistics in one place.
Statistic functions inside **decoupleR** always return a tidy tibble that can be
easily processed with the tools provide by the [tidyverse ecosystem](https://www.tidyverse.org/).

```{r usage-decouple_function, message=FALSE}
decouple(
    mat = mat,
    network = network,
    .source = "tf",
    .target = "target",
    statistics = c("gsva", "wmean", "wsum", "ulm", "ora"),
    args = list(
        gsva = list(verbose = FALSE),
        wmean = list(times=100),
        wsum = list(times=100),
        ulm = list(center=FALSE),
        ora = list()
    )
)
```

### Individual parts

In turn, we recognize that the use of individual statistics may be of interest.
Therefore, these are also exported and ready for use. All statistics follow the
same design pattern and arguments, so moving between statistics could be very comfortable.

```{r}
# wmean call is equivalent to the one made by decouple() above.
run_wmean(
    mat = mat,
    network = network,
    .source = "tf",
    .target = "target",
    .likelihood = "likelihood"
)
```

<!-- ## Citation -->

<!-- Below is the citation output from using `citation('decoupleR')` in R. Please -->
<!-- run this yourself to check for any updates on how to cite __decoupleR__. -->

<!-- ```{r 'citation', eval = requireNamespace('decoupleR')} -->
<!-- print(citation("decoupleR"), bibtex = TRUE) -->
<!-- ``` -->

<!-- Please note that the `decoupleR` was only made possible thanks to many other R -->
<!-- and bioinformatics software authors, which are cited either in the vignettes -->
<!-- and/or the paper(s) describing this package. -->

## Contributing to decoupleR

Are you interested in adding a new statistical method or collaborating in the
development of internal tools that allow the extension of the package?
Please check out our
[contribution guide](https://saezlab.github.io/decoupleR/CONTRIBUTING.html).

---

Please note that this project is released with a [Contributor Code of Conduct](https://saezlab.github.io/decoupleR/CODE_OF_CONDUCT).
By participating in this project you agree to abide by its terms.
