---
title: "Introduction to decoupleR"
author:
  - name: Jesús Vélez
    affiliation:
    - National Autonomous University of Mexico
    email: jvelezmagic@gmail.com
output: 
  BiocStyle::html_document:
    self_contained: true
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
package: "`r pkg_ver('decoupleR')`"
vignette: >
  %\VignetteIndexEntry{Introduction to decoupleR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r chunk_setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

```{r vignette_setup, echo=FALSE, message=FALSE, warning = FALSE}
# Track time spent on making the vignette.
start_time <- Sys.time()

# Bib setup.
library(RefManageR)

# Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    RefManageR = citation("RefManageR")[1],
    decoupleR = citation("decoupleR")[1],
    GSVA = citation("GSVA")[1],
    viper = citation("viper")[1]
)
```

# Basics

## Install `decoupleR`

`R` is an open-source statistical environment which can be easily modified to
enhance its functionality via packages. `r Biocpkg("decoupleR")` is an `R`
package available via the [Bioconductor](http://bioconductor.org) repository 
for packages. `R` can be installed on any operating system from
[CRAN](https://cran.r-project.org/) after which you can install
`r Biocpkg("decoupleR")` by using the following commands in your `R` session:

```{r bioconductor_install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("decoupleR")

# Check that you have a valid Bioconductor installation
BiocManager::valid()
```

You can install the development version from [GitHub](https://github.com/) with:

```{r github_install, eval=FALSE}
BiocManager::install("saezlab/decoupleR")
```

## Required knowledge

`r Biocpkg("decoupleR")` is based on many other packages and in particular in 
those that have implemented the infrastructure needed for dealing with
functional analysis. That is, packages like `r Biocpkg("viper")` or 
`r Biocpkg("GSVA")`, among others. This in order to have a centralized place
from which to apply different statistics to the same data set without the need
and work that would require testing in isolation. Opening the possibility of
developing benchmarks that can grow easily.

## Asking for help

As package developers, we try to explain clearly how to use our packages and in
which order to use the functions. But `R` and `Bioconductor` have a steep
learning curve so it is critical to learn where to ask for help.
We would like to highlight the
[Bioconductor support site](https://support.bioconductor.org/) as the main
resource for getting help: remember to use the `decoupleR` tag and check 
[the older posts](https://support.bioconductor.org/t/decoupleR/).
Other alternatives are available such as creating GitHub issues and tweeting.
However, please note that if you want to receive help you should adhere to the
[posting guidelines](http://www.bioconductor.org/help/support/posting-guide/).
It is particularly critical that you provide a small reproducible example and
your session information so package developers can track down the source of
the error.

## Citing `decoupleR`

We hope that `r Biocpkg("decoupleR")` will be useful for your research.
Please use the following information to cite the package and the overall
approach. Thank you!

```{r decoupleR citation}
citation("decoupleR")
```


# Quick start to using to `decoupleR`

## Libraries

`r Biocpkg("decoupleR")` provides different statistics to calculate the
regulatory activity given a `matrix` of molecular readouts and a `network`.
It incorporates pre-existing methods to avoid recreating the wheel while
implementing its own methods under an evaluation standard.
Therefore, it provides flexibility when evaluating a data set with different
statistics.
  
Since inputs and outputs are always tibbles (i.e. special data frames),
incorporating `r Githubpkg("tidyverse/dplyr","dplyr")` into your workflow can
be useful for manipulating results, but it is not necessary.

```{r load_library, message=FALSE}
library(decoupleR)
library(dplyr)
```

## Input data

In order to use it, you first need to have a `matrix` where the rows represent
the target nodes and the columns the samples. In addition, it is necessary to 
provide a prior knowledge `network` that contains at least two columns corresponding to the 
source and target nodes. It is noteworthy that certain methods will require 
specifying additional metadata columns. For instance, the mode of regulation 
(MoR) or the likelihood of the interaction.

```{r read_example_data}
inputs_dir <- system.file("testdata", "inputs", package = "decoupleR")

mat <- file.path(inputs_dir, "input-expr_matrix.rds") %>%
    readRDS() %>%
    glimpse()

network <- file.path(inputs_dir, "input-dorothea_genesets.rds") %>%
    readRDS() %>%
    glimpse()

# We recommend to intersect and set a minsize of targets
network <- intersect_regulons(mat, network, tf, target, minsize=5)
```

## How to decouple?

Once the data is loaded, you are one step away from achieving decoupling.
This step corresponds to specifying which statistics you want to run.
For more information about the defined statistics and their parameters,
you can execute `?decouple()`.

```{r usage-decouple_function, message=TRUE}
decouple(
    mat = mat,
    network = network,
    .source = "tf",
    .target = "target",
    statistics = c("gsva", "wmean", "mlm", "ora"),
    args = list(
        gsva = list(verbose = FALSE),
        wmean = list(times = 100),
        mlm = list(center=FALSE),
        ora = list(n_up=300, n_bottom=300, n_background=20000)
    )
) %>%
    glimpse()
```

Done, you have applied different statistics to the same data set, now you can
analyze them at your convenience, for example, by performing a [benchmark]().

## How it works?

### Mapping network columns

To carry out the column mapping, `decoupleR` relies on the selection provided
by the `r Githubpkg("tidyverse/tidyselect","tidyselect")` package.
Some of the selection it provides are:

- Symbols
- Strings
- Position

Let's see an example. Input network has the following columns:
```{r show_columns}
network %>%
    colnames()
```

We can use the way we like to do the mapping, even a combination of ways to do
it. This applies not only to the decouple function, but to all functions of
the `decoupleR statistics` family, identifiable by the `run_` prefix.

```{r}
this_column <- "target"
viper_res <- decouple(
    mat = mat,
    network = network,
    .source = tf,
    .target = !!this_column,
    statistics = c("viper"),
    args = list(
        viper = list(
            .mor = 4,
            .likelihood = "likelihood",
            verbose = FALSE
        )
    ),
    show_toy_call = TRUE
)
```

# Reproducibility

## Special thanks

The `r Biocpkg("decouopleR")` package `r Citep(bib[["decoupleR"]])` was made
possible thanks to:

* R `r Citep(bib[["R"]])`
* `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
* `r CRANpkg("knitcitations")` `r Citep(bib[["knitcitations"]])`
* `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])`
* `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])`
* `r CRANpkg("sessioninfo")` `r Citep(bib[["sessioninfo"]])`
* `r CRANpkg("testthat")` `r Citep(bib[["testthat"]])`
* `r Biocpkg("GSVA")` `r Citep(bib[["GSVA"]])`
* `r Biocpkg("viper")` `r Citep(bib[["viper"]])`

This package was developed using
`r BiocStyle::Githubpkg("lcolladotor/biocthis")`.

## Vignette

### Create

```{r create_vignette, eval=FALSE}
# Create the vignette
library(rmarkdown)
system.time(render("decoupleR.Rmd", "BiocStyle::html_document"))

# Extract the R code
library(knitr)
knit("decoupleR.Rmd", tangle = TRUE)
```

### Wallclock time spent generating the vignette

```{r reproduce_time, echo=FALSE}
# Processing time in seconds
total_time <- diff(c(start_time, Sys.time()))
round(total_time, digits = 3)
```

## Session information

```{r session_info, echo=FALSE}
options(width = 120)
sessioninfo::session_info()
```

# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` 
`r Citep(bib[["BiocStyle"]])` with `r CRANpkg("knitr")`
`r Citep(bib[["knitr"]])` and `r CRANpkg("rmarkdown")`
`r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg('RefManageR')` `r Citep(bib[['RefManageR']])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```
